Multi-state modelling with R: the {\tt msm} package \vskip 0.2in \large{Version <<echo=FALSE,results=tex>>= cat(version) @ \vskip 0.1in <<echo=FALSE,results=tex>>= cat(format(Sys.time(), " @ }

### 2014-07-08 src/doc/msm-manual.Rnw
version 1.4

<<echo=FALSE>>=
version <- gsub("Version: +", "",
               packageDescription("msm", lib.loc="../../..")$Version)
@

@@

Figure~\ref{fig:multi} illustrates a multi-state model in continuous
time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
the individual is in state $S(t)$.  The arrows show which transitions
are possible between states. The next state to which the individual
moves, and the time of the change, are governed by a set of {\em
  transition intensities} $q_{rs}(t, z(t))$ for each pair of states
$r$ and $s$.  The intensities may also depend on the time of the
process $t$, or more generally a set of individual-specific or
time-varying explanatory variables $z(t)$.  The intensity represents
the instantaneous risk of moving from state $r$ to state $s$:
\begin{equation}
  \label{eq:multi:intensity}
  q_{rs}(t, z(t))  =  \lim_{\delta t \rightarrow 0} P (S(t+\delta t) = s | S(t) = r) / \delta t
\end{equation}
The intensities form a matrix $Q$ whose rows sum to zero, so that the
diagonal entries are defined by $q_{rr} = - \sum_{s \neq r} q_{rs}$.

@@

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
1 + \frac{(-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models, illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
structure on any number of states in principle, $P(t)$ is
determined from the matrix exponential.  This is calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

<<>>=
twoway4.q  <-  rbind ( c(0, 0.25, 0, 0.25),
                       c(0.166, 0, 0.166, 0.166),
                       c(0, 0.25, 0, 0.25),
                       c(0, 0, 0, 0) )
@

@@

Such default initial values can be used by supplying
\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm} 
could be used to get the  amatrix of initial values as follows.  
These methods are only available for non-hidden Markov models.

<<>>=
twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, 
                                   qmatrix=twoway4.q)
@

However, if there are are many changes of state in between the
observation times, then this crude approach may fail to give sensible
initial values.  For the heart transplant example we could also guess
that the mean period in each state before moving to the next is about
2 years, and there is an equal probability of progression, recovery or
death.  This gives $q_{rr} = - 0.5$ for $r = 1, 2, 3$, and $q_{12} =
q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
0.25$, and the initial value matrix \Robject{twoway4.q} shown above,
which we now use to fit the model.

@@

To fit the model, call the \Rfunction{msm} function with the appropriate
arguments.  For our running example, we have defined a data set
\Robject{cav}, a matrix \Robject{twoway4.q} indicating the allowed
transitions, and initial values. We are ready to run \Rfunction{msm}.

@@

<<>>=
cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                    qmatrix = twoway4.q, death = 4 )
@

@@

<<eval=FALSE>>=
help(optim)
@
for more options to control the optimisation.  When completed, the
\Rfunction{msm} function returns a value.  This value is a list of the
important results of the model fitting, including the parameter
estimates and their covariances.  To keep these results for
post-processing, we store them in an R object, here called
\Robject{cav.msm}.  When running several similar \Rfunction{msm} models,
it is recommended to store the respective results in
informatively-named objects.

@@

From the estimated intensity matrix, we see patients are three times
as likely to develop symptoms than die without symptoms (first row).
After disease onset (state 2), progression to severe symptoms (state
3) is 50\% more likely than recovery.  Once in the severe state, death
is more likely than recovery, and a mean of 1 / -0.44 = 2.3 years is
spent in state 3 before death or recovery.

@@

\paragraph{Model 2: sex as a covariate}
<<>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                     qmatrix = twoway4.q, death = 4, covariates = ~ sex)
@

The \Robject{msm} object will now include the estimated covariate
effects and their confidence intervals.

<<>>=
cavsex.msm
@
Comparing the estimated log-linear effects of sex and their standard
errors, we see that the disease onset rate is smaller for females,
whereas none of the other effects are significant. The first matrix shown in the output of printing
\Robject{cavsex.msm} is the estimated transition intensity matrix
$q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
set to its mean value in the data.  This represents an average
intensity matrix for the population of 535 male and 87 female
patients. To extract separate intensity matrices for male and female
patients ($z = 0$ and $1$ respectively), use the function
\Rfunction{qmatrix.msm}, as shown below.  This and similar summary
functions will be described in more detail in
section \ref{sec:extractor}.

@@

Different transition rates may be easily modelled on different
covariates by specifying a named list of formulae as the
\Rfunarg{covariates} argument.  Each element of the list has a name
identifying the transition.  In the model below, the transition rate
from state 1 to state 2 and the rate from state 3 to state 4 are each
modelled on sex as a covariate, but no other intensities have
covariates on them.

\paragraph{Model 2a: transition-specific covariates}
<<eval=FALSE>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                  qmatrix = twoway4.q, death = 4,
                  covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
@

@@

<<eval=FALSE>>=
cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                qmatrix = twoway4.q, death = 4,
                covariates = ~ sex, 
                constraint = list(sex=c(1,2,3,1,2,3,2)) )
@

@@

\paragraph{Model 4: fixed parameters}
<<eval=FALSE>>=
cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                qmatrix = twoway4.q, death = 4,
                control = list(trace=2, REPORT=1),
                fixedpars = c(6, 7) )
@

A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
effect parameters to zero, that is to assume no effect of a covariate
on a certain transition rate.

@@

@@

\item[Hazard ratios for transition] The function \Rfunction{hazard.msm}
  gives the estimated hazard ratios corresponding to each covariate
  effect on the transition intensities. 95\% confidence limits are
  computed by assuming normality of the log-effect. For example, for
  model 2 with female sex as a covariate, the following hazard ratios
  show more clearly that the only transition on which the effect of
  sex is significant at the 5\% level is the 1-2 transition.

@@

For most output extractor functions provided with \Rpackage{msm}, the
option \Rfunarg{ci=''boot''} is available, as a wrapper around
\Rfunction{boot.msm}, to enable bootstrap confidence intervals to be calculated.
But any user-defined output statistic can be bootstrapped, as follows.
The function \Rfunction{boot.msm} is called with the fitted
\Rfunction{msm} model as first argument, and an R function specifying
the statistic to be bootstrapped as the second argument \texttt{stat}.
The return value from \Rfunction{boot.msm} is a list of \texttt{B}
replicates (by default, \texttt{B=1000}) of the desired statistic.  For
example, to bootstrap the transition intensity matrix of the heart
transplantation model \Robject{cav.msm}:
\begin{Scode}
  q.list <- boot.msm(cav.msm, stat=function(x){qmatrix.msm(x)$estimates})
\end{Scode}
Note that for \Rfunction{boot.msm} to be able to refit the original
model that produced \Robject{cav.msm}, all objects used in the
original model fit (for example, in this case, \Robject{twoway4.q})
must be in the working environment.  Otherwise, \Rfunction{boot.msm}
will give an ``object not found'' error.

@@

\begin{description}
\item[Initial values] Make sure that a sensible set of initial values
  have been chosen. The optimisation may only converge within a
  limited range of `informative' initial values.  It is also sensible
  to run the model for several different initial values to ensure that
  the estimation has converged to a global rather than a local optimum.

@@

\item[Choice of algorithm] By default, \Rfunction{msm} uses a
  Nelder-Mead algorithm which does not use analytic derivatives of the
  objective function.   For potentially greater speed and accuracy,
  consider using a quasi-Newton algorithm such as
  the ``BFGS'' method of \Rfunction{optim}, which can make use of
  analytic derivatives, for example, \Rfunarg{method = ``BFGS'', use.deriv = TRUE}.
  Or use the \Rfunction{nlm} R function via \Rfunarg{msm(..., opt.method = "nlm" , ...)}
  Note also the Fisher scoring method available for non-hidden Markov
  models for panel data, via  \Rfunarg{msm(..., opt.method = "fisher", ...)},
  is expected to be faster than the generic methods, but less robust to
  bad initial values.

@@

  If none of these numerical adjustments lead to convergence, then the
  model is probably over-complicated.  There may not be enough
  information in the data on a certain transition rate.  It is always
  recommended to count all the pairs of transitions between states in
  successive observation times, making a frequency table of previous
  state against current state (function \Rfunction{statetable.msm}).
  Although the data are a series of snapshots of a continuous-time
  process, and the actual transitions take place in between the
  observation times, this type of table may still be helpful.  If
  there are not many observed `transitions' from state 2 to state 4,
  for example, then the data may be insufficient to estimate $q_{24}$.

@@

To model observed states with misclassification, we define an matrix
\Rfunarg{ematrix} indicating the states that can be misclassified.
Rows of this matrix correspond to true states, columns to observed
states.  It should contains zeroes in the positions where
misclassification is not permitted.  Non-zero entries are initial
values for the corresponding misclassification probabilities. We then
call \Rfunction{msm} as before, but with this matrix as the
\Rfunarg{ematrix} argument.  Initial values of 0.1 are assumed for
each of the four misclassification probabilities $e_{12}, e_{21},
e_{23}, e_{32}$.  Zeroes are given where the elements of $E$ are zero.
The diagonal elements supplied in \Rfunarg{ematrix} are ignored, as
rows must sum to one. The matrix \Rfunarg{qmatrix}, specifying
permitted transition intensities and their initial values, also
changes to correspond to the new $Q$ representing the progression-only
model for the underlying states.

@@

<<>>=
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                     c(0, 0, 0.202, 0.081),
                     c(0, 0, 0, 0.126),
                     c(0, 0, 0, 0))
ematrix <- rbind(c(0, 0.1, 0, 0),
                 c(0.1, 0, 0.1, 0),
                 c(0, 0.1, 0, 0),
                 c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                   qmatrix = oneway4.q, ematrix = ematrix, death = 4,
                   obstrue = firstobs, method="BFGS")
cavmisc.msm
@

@@

<<>>=
cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                      qmatrix = oneway4.q, ematrix = ematrix, 
                      death = 4, misccovariates = ~sex, 
                      obstrue=firstobs, method="BFGS")
@
<<>>=
cavmiscsex.msm
@
The large confidence interval for the 1/2 misclassification suggests
that women are no more or less likely than men to have a false
positive angiography scan. On the other hand, women have slightly more
false negatives.

@@

<<>>=
ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
@
The confidence intervals for the estimates for women are higher, since
there are only 87 women in this set of 622 patients.

@@

fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
                death=3, hmodel=hmodel1,
                hcovariates=list(~acute, ~acute, NULL),
                hcovinits = list(-8, -8, NULL),
                hconstraint = list(acute = c(1,1)), method="BFGS")

@@

Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
and confidence intervals for the transition intensity matrix and the
hidden Markov model parameters.  The estimated within-state means of
FEV$_1$ are around 98\% and 52\% baseline respectively.  From the
estimated transition intensities, individuals spend around 1421 days
(3.9 years) before getting BOS, after which they live for an average
of 1248 days (3.4 years).  FEV$_1$ is lower by an average of 8\%
baseline within 14 days of acute events.

@@

\begin{Scode}
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                   c(0, 0, 0.202, 0.081),
                   c(0, 0, 0, 0.126),
                   c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), 
                                    hmmCat(c(0.1, 0.8, 0.1, 0)), 
                                    hmmCat(c(0, 0.1, 0.9, 0)), 
                                    hmmIdent(4)),
                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, 
                     method="BFGS")
cavmisc.msm
\end{Scode}

@@

>>>>

<<echo=FALSE>>=
version <- gsub("Version: +", "",
               packageDescription("msm", lib.loc=c("../..",.libPaths()))$Version)
@

@@

Figure \ref{fig:multi} illustrates a multi-state model in continuous
time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
the individual is in state $S(t)$.  The arrows show which transitions
are possible between states. The next state to which the individual
moves, and the time of the change, are governed by a set of {\em
  transition intensities} $q_{rs}(t, z(t))$ for each pair of states
$r$ and $s$.  The intensities may also depend on the time of the
process $t$, or more generally a set of individual-specific or
time-varying explanatory variables $z(t)$.  The intensity represents
the instantaneous risk of moving from state $r$ to state $s$:
\begin{equation}
  \label{eq:multi:intensity}
  q_{rs}(t, z(t))  =  \lim_{\delta t \rightarrow 0} P (S(t+\delta t) = s | S(t) = r) / \delta t
\end{equation}
The intensities form a matrix $Q$ whose rows sum to zero, so that the
diagonal entries are defined by $q_{rr} = - \sum_{s \neq r} q_{rs}$.

@@

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t})  & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
1 - e^{-(q_{12} + q_{13})t} - \frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t}) & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models, illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
structure on any number of states in principle, $P(t)$ is
determined from the matrix exponential.  This is calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

<<>>=
Q  <-  rbind ( c(0, 0.25, 0, 0.25),
               c(0.166, 0, 0.166, 0.166),
               c(0, 0.25, 0, 0.25),
               c(0, 0, 0, 0) )
@

@@

Such default initial values can be used by supplying
\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm}
could be used to get this matrix of initial values explicitly as follows.
These methods are only available for non-hidden Markov models.

<<>>=
Q.crude  <- crudeinits.msm(state ~ years, PTNUM, data=cav,
                                   qmatrix=Q)
@

However, if there are are many changes of state in between the
observation times, then this crude approach may fail to give sensible
initial values.  For the heart transplant example we could also guess
that the mean period in each state before moving to the next is about
2 years, and there is an equal probability of progression, recovery or
death.  This gives $q_{rr} = - 0.5$ for $r = 1, 2, 3$, and $q_{12} =
q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
0.25$, and the initial value matrix \Robject{Q} shown above,
which we now use to fit the model.

@@

To fit the model, call the \Rfunction{msm} function with the appropriate
arguments.  For our running example, we have defined a data set
\Robject{cav}, a matrix \Robject{Q} indicating the allowed
transitions, and initial values. We are ready to run \Rfunction{msm}.

@@

<<>>=
cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                    qmatrix = Q, death = 4)
@

@@

<<eval=FALSE>>=
help(optim)
@
for more options to control the optimisation. \footnote{Note that since version 1.3.2, \Rfunarg{method=''BFGS''}, is the default optimisation algorithm in \Rfunction{msm}, since it can use analytic derivatives, which are available for most models.}

When completed, the \Rfunction{msm} function returns a value.  This
value is a list of the important results of the model fitting,
including the parameter estimates and their covariances.  To keep
these results for post-processing, we store them in an R object, here
called \Robject{cav.msm}.  When running several similar
\Rfunction{msm} models, it is recommended to store the respective
results in informatively-named objects.

@@

From the estimated intensities, we see patients are three times
as likely to develop symptoms than die without symptoms (transitions from state 1).
After disease onset (state 2), progression to severe symptoms (state
3) is 50\% more likely than recovery.  Once in the severe state, death
is more likely than recovery, and a mean of 1 / -0.44 = 2.3 years is
spent in state 3 before death or recovery.

@@

\paragraph{Model 2: sex as a covariate}
<<>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                     qmatrix = Q, death = 4, covariates = ~ sex)
@

Printing the \Robject{msm} object now displays the estimated covariate
effects and their confidence intervals (note since version 1.3.2 these
are \emph{hazard ratios} $\exp(\beta_{rs})$, not \emph{log hazard ratios} $\beta_{rs}$ as in previous
versions).

<<>>=
cavsex.msm
@
The sizes of the
confidence intervals for some of the hazard ratios suggests there is no information in the data about 
the corresponding covariate effects, leading to a likelihood that is a flat function of these parameters, 
and this model should be simplified. The first column shown in the output
is the estimated transition intensity matrix
$q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
set to its mean value in the data.  This represents an average
intensity matrix for the population of 535 male and 87 female
patients. To extract separate intensity matrices for male and female
patients ($z = 0$ and $1$ respectively), use the function
\Rfunction{qmatrix.msm}, as shown below.  This and similar summary
functions will be described in more detail in
section \ref{sec:extractor}.

@@

Since \Rpackage{msm} version 1.2.3, different transition rates may be
easily modelled on different covariates by specifying a named list of
formulae as the \Rfunarg{covariates} argument.  Each element of the
list has a name identifying the transition.  In the model below, the
transition rate from state 1 to state 2 and the rate from state 1 to
state 4 are each modelled on sex as a covariate, but no other
intensities have covariates on them.

\paragraph{Model 2a: transition-specific covariates}
<<eval=FALSE>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                  qmatrix = Q, death = 4,
                  covariates = list("1-2" = ~ sex, "1-4" = ~sex) )
@

@@

<<eval=FALSE>>=
cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                qmatrix = Q, death = 4,
                covariates = ~ sex,
                constraint = list(sex=c(1,2,3,1,2,3,2)) )
@

@@

\paragraph{Model 4: fixed parameters}
<<eval=FALSE>>=
cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                qmatrix = Q, death = 4,
                control = list(trace=2, REPORT=1),
                fixedpars = c(6, 7) )
@

A \Rfunarg{fixedpars} statement can also be used for fixing covariate
effect parameters to zero, that is to assume no effect of a covariate
on a certain transition rate.

@@

\item[Expected number of visits] The function \Rfunction{envisits.msm}
  estimates the expected number of visits to a state, computed in a
  similar way to the total length of stay. See its help page for
  further details.

@@

\item[Hazard ratios for transition] The function \Rfunction{hazard.msm}
  gives the estimated hazard ratios corresponding to each covariate
  effect on the transition intensities. 95\% confidence limits are
  computed by assuming normality of the log-effect.

@@

For most output extractor functions provided with \Rpackage{msm}, the
option \Rfunarg{ci=''boot''} is available, as a wrapper around
\Rfunction{boot.msm}, to enable bootstrap confidence intervals to be calculated.
But any user-defined output statistic can be bootstrapped, as follows.
The function \Rfunction{boot.msm} is called with the fitted
\Rfunction{msm} model as first argument, and an R function specifying
the statistic to be bootstrapped as the second argument \texttt{stat}.
The return value from \Rfunction{boot.msm} is a list of \texttt{B}
replicates (by default, \texttt{B=1000}) of the desired statistic.  For
example, to bootstrap the transition intensity matrix of the heart
transplantation model \Robject{cav.msm}:
\begin{Scode}
  q.list <- boot.msm(cav.msm, stat=function(x){qmatrix.msm(x)$estimates})
\end{Scode}
Note that for \Rfunction{boot.msm} to be able to refit the original
model that produced \Robject{cav.msm}, all objects used in the
original model fit (for example, in this case, \Robject{Q})
must be in the working environment.  Otherwise, \Rfunction{boot.msm}
will give an ``object not found'' error.

@@

  Firstly, make sure there are not too many parameters in the model.  There may not be enough
  information in the data on a certain transition rate.  It is
  recommended to count all the pairs of transitions between states in
  successive observation times, making a frequency table of previous
  state against current state (function \Rfunction{statetable.msm}),
  and do this for any subgroups defining covariates.
  Although the data are a series of snapshots of a continuous-time
  process, and the actual transitions take place in between the
  observation times, this type of table may still be helpful.  If
  there are not many observed `transitions' from state 2 to state 4,
  for example, then the data may be insufficient to estimate $q_{24}$.

\item[Initial values] Make sure that a sensible set of initial values
  have been chosen. The optimisation may only converge within a
  limited range of `informative' initial values.  It is also sensible
  to run the model for several different initial values to ensure that
  the estimation has converged to a global rather than a local optimum.

@@

\item[Choice of algorithm] By default, since version 1.3.2,
  \Rfunction{msm} uses the BFGS method of \Rfunction{optim}, which
  makes use of analytic derivatives.  Analytic derivatives are
  available for all models in msm, apart from hidden Markov models
  with unknown initial state probabilities, misclassification models
  with equality constraints on misclassification probabilities, and
  truncated or measurement-error outcome distributions.  This speeds
  up optimisation.  Though alternative algorithms are available such
  as \Rfunarg{method = ``CG''}.  Or use the \Rfunction{nlm} R function
  via \Rfunarg{msm(..., opt.method = "nlm" , ...)}  Note also the
  Fisher scoring method available for non-hidden Markov models for
  panel data, via \Rfunarg{msm(..., opt.method = "fisher", ...)}, is
  expected to be faster than the generic methods, but less robust to
  bad initial values.  Or since version 1.3.2, msm can also use
  \Rfunarg{method=``bobyqa''} from the \Rpackage{minqa} package, a fast
  derivative-free method.

@@

@@

To model observed states with misclassification, we define a matrix
\Rfunarg{ematrix} indicating the states that can be misclassified.
Rows of this matrix correspond to true states, columns to observed
states.  It should contains zeroes in the positions where
misclassification is not permitted.  Non-zero entries are initial
values for the corresponding misclassification probabilities. We then
call \Rfunction{msm} as before, but with this matrix as the
\Rfunarg{ematrix} argument.  Initial values of 0.1 are assumed for
each of the four misclassification probabilities $e_{12}, e_{21},
e_{23}, e_{32}$.  Zeroes are given where the elements of $E$ are zero.
The diagonal elements supplied in \Rfunarg{ematrix} are ignored, as
rows must sum to one. The matrix \Rfunarg{qmatrix}, specifying
permitted transition intensities and their initial values, also
changes to correspond to the new $Q$ representing the progression-only
model for the underlying states.

@@

<<>>=
Qm <- rbind(c(0, 0.148, 0, 0.0171),
            c(0, 0, 0.202, 0.081),
            c(0, 0, 0, 0.126),
            c(0, 0, 0, 0))
ematrix <- rbind(c(0, 0.1, 0, 0),
                 c(0.1, 0, 0.1, 0),
                 c(0, 0.1, 0, 0),
                 c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                   qmatrix = Qm, ematrix = ematrix, death = 4,
                   obstrue = firstobs)
cavmisc.msm
@

@@

<<>>=
cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                      qmatrix = Qm, ematrix = ematrix,
                      death = 4, misccovariates = ~sex,
                      obstrue=firstobs)
@
<<>>=
cavmiscsex.msm
@
The large confidence interval for the odds ratio for 1/2 misclassification suggests
there is no information in the data about the difference between genders in 
the false positive rates for angiography. On the other hand, women have slightly more
false negatives.

@@

<<>>=
ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
@
The confidence intervals for the estimates for women are wider, since
there are only 87 women in this set of 622 patients.

@@

fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
                death=3, hmodel=hmodel1,
                hcovariates=list(~acute, ~acute, NULL),
                hcovinits = list(-8, -8, NULL),
                hconstraint = list(acute = c(1,1)))

@@

Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
and confidence intervals for the transition intensities and the
hidden Markov model parameters.  The estimated within-state means of
FEV$_1$ are around 98\% and 52\% baseline respectively.  From the
estimated transition intensities, individuals spend around 1421 days
(3.9 years) before getting BOS, after which they live for an average
of 1248 days (3.4 years).  FEV$_1$ is lower by an average of 8\%
baseline within 14 days of acute events.

@@

\begin{Scode}
Qm <- rbind(c(0, 0.148, 0, 0.0171),
            c(0, 0, 0.202, 0.081),
            c(0, 0, 0, 0.126),
            c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)),
                                    hmmCat(c(0.1, 0.8, 0.1, 0)),
                                    hmmCat(c(0, 0.1, 0.9, 0)),
                                    hmmIdent(4)),
                     qmatrix = Qm, obstrue=firstobs, death = 4)
cavmisc.msm
\end{Scode}

@@

\item (Optionally) Write a C function to compute the derivatives of
  the probability density with respect to the parameters, and put this
  in \texttt{src/hmmderiv.c}, with a declaration in
  \texttt{src/hmm.h}.  Then add the model to \texttt{DHMODELS} in
  \texttt{lik.c} (in the same position as in \texttt{HMODELS}) and
  \texttt{.msm.HMODELS.DERIV} in \texttt{R/constants.R}.  This will
  generally double the speed of maximum likelihood estimation, but analytic 
  derivatives will not be available for all distributions.
